Assessment and implementation of resource road simulation methods for projection of cumulative effects on wildlife

### 2023-01-10 analysis/paper/paper.Rmd
writing in progress

---
title: "Assessment and implementation of forestry road network projection methods"
author:
  - name: Josie Hughes
    email: josie.hughes@ec.gc.ca
    institute: [ECCC]
    correspondence: true
  - name: Sarah Endicott
    email: sarah.endicott@ec.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: David Lapins
    email: david.lapins@pc.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Kyle Lochhead
    email: Kyle.Lochhead@gov.bc.ca
    institute: [FLNR]
    correspondence: false
  - name: Gregory Paradis
    email: gregory.paradis@ubc.ca
    institute: [UBC]
    correspondence: false

@@

The impacts of roads on wildlife are complex and varied. Mortality risk and stress on or near roads can be elevated due to vehicle collisions, hunting pressure, predation, noise and other stressors, but roads can also provide food, facilitate movement, and provide refuge [@hill_a_2021;@trombulak_review_2000;@fahrig_effects_2009].    
In Canadian forests, there is ample evidence of linear features attracting predators such as black bears and wolves seeking food or easy movement, while repelling prey such as caribou and moose [@dickie_corridors_2020;@polfus_identifying_2011;@mumma_predation_2018;@beauchesne_disentangling_2013;@leblond_avoidance_2013;@laurian_interactions_2012]; there is also evidence of predators avoiding features heavily used by humans [@muhly_human_2011;@whittington_towns_2022]. Changes in mortality risks, habitat quality, and connectivity associated with anthropogenic disturbance can in turn alter movement patterns [@tucker_moving_2018], demography and population viability [@johnson_science_2020; @decesare_linking_2014; @barrientos_the_2021;@apps_spatial_2013;@pinard_calving_2012;@fahrig_effects_2009]. Roads can also facilitate spread of invasive species [], and enable other development activities [@johnson_growth-inducing_2020].

Increasingly rapid environmental change, improvements in available cyberinfrastructure, and development of open source data, tools and methods have contributed to growing interest in ecological forecasting models [bodner, dietze, perfict]. Recent efforts to forecast impacts of climate change and anthropogenic disturbance on Canadian forests and forest-dwelling wildlife [@tremblay_harvesting_2018;@snell_importance_2018;@daniel_incorporating_2017;@stewart_;@michelletti_;recent quebec examples] build on a rich history of spatial-stochastic models of forest growth and succession following disturbance [landis;overview;stms;seles]. Despite the importance of roads for wildlife, none of these efforts have included forecasts of the development of road networks associated with forest harvest, largely because tools to project road network development over large spatial scales have not been available.

Forest roads are also essential for harvest planning. Some forest management decision support systems include forest road projection methods [patchworks], but methods implemented in proprietary closed-source commercial software cannot be used in other contexts. Published methods intended to support strategic planning [@Anderson_projecting_2004;Clark et al., 2000; Dean, 1997] require inputs that are not publicly available, and employ algorithms that are too complex and detailed for application at regional, provincial or national scales. More promising options for projecting forest road network development at large scales include an iterative least cost path method developed by Clement [2021], and a least cost path minimum spanning tree algorithm used to project roads and reconstruct development history in British Columbia [lochhead published and not]. The former method is available as QGIS and Landis plugins, and the latter is embedded in a larger open source modelling project that is specific to British Columbia []. Neither method has been formally assessed, compared, published, or implemented in a way that allows for easy integration into the variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [spades/syncrosim/landis].

<!-- Comment on metrics relevant to wildlife and what exactly accuracy means. -->
In this work, we address the need for adequately accurate forest road network projection tools that are also widely and easily applicable by implementing simple iterative least cost path and minimum spanning tree methods in an open source R package. We compare the ability of these methods to reconstruct observed forest road network development since 1990 in a mountainous region of British Columbia using performance metrics relevant to wildlife.

@@

<!-- TO DO: dynamic links to figs and tables. -->
We chose the Revelstoke timber supply area (TSA 27), in British Columbia, Canada as an assessment area because forestry is the dominant resource extraction activity in this region and much of the road network was built to access to forest harvest cutblocks (Figure 1). This allows projection of the forest road network from historic forest harvest data. Approximately 55% of the area is non-forested land where road building is difficult or impossible (e.g. alpine, lakes, swamp, brush, rock) [@brown2010], so the area provides opportunities to observed the responses of road projection algorithms to barriers. Revelstoke is in the interior-wet belt of the province where forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%) [@ministryofforests]. The area contains wildlife species that are sensitive to roads (refs) including endangered southern mountain woodland caribou (refs).

@@

Our road network data was obtained by combining the digital road atlas (DRA) [@databc2021] and forest tenure road [@britishcolumbiaministryofforestslandsnaturalresourceoperationsandruraldevelopmentflnrord2021] data sets for BC. To avoid situations where the same road was included in both data sets with slightly different coordinates, we removed all roads from the DRA data set that were within 50 m of a road in the forest tenure road data set. The existing road network, used as the starting point for projections, was created by filtering out forest tenure roads with an award date greater than 1990. All roads from the DRA data set were assumed to be built before 1990. Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset [@governmentofbritishcolumbia2021b]. Cutblock information is compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes. The dataset does not exclude non-harvested 'reserved' areas, and satellite image change detection may not always represent the entire cutblock area [@governmentofbritishcolumbia2021b]. Forest harvest cutblocks with harvest years greater than 1990 were filtered and used as targets for the projections.

To spatially define the cost of building and traveling on a forestry road we developed a simple model informed by slope and major barriers (Table 1). Following the Interior Appraisal Manual of BC, we assume the base cost per kilometer of road is a linear function of slope [@governmentofbritishcolumbia2021a], and that road construction across major water bodies is prohibitively expensive (Table 1). Existing roads are assigned a cost of 0. The cost raster consists of 1703936 pixels with a resolution of 100 meters by 100 meters.

@@

All data sets were accessed using the `bcdata` R package and filtered to the extent of the Revelstoke TSA [@teucher2021].

@@

Simple methods for projecting low-cost routes from an existing road network to target locations (i.e. harvest cutblocks) include simple least cost paths (LCP), dynamic least cost paths (DLCP) and least cost path minimum spanning trees (MST) \@ref(fig:fig2)). We use Dijkstra's shortest path algorithm [-@dijkstra1959] to determine the least cost path between a target location (e.g. centroid of a cutblock) and the existing road network. Least cost paths to each cutblock is determined independently in simple method (LCP). The dynamic least cost paths (DLCP) method evaluates the path to each cutblock in sequence so that earlier paths are considered in the cost for later paths. Sequence is determined by a heuristic such as the distance to the nearest existing road in these examples. The minimum spanning tree (MST) method calculates the minimum cost network to connecting all cutblocks to the road network using Kruskal's algorithm [-@kruskal1956]. See Appendix A for a more detailed explanation of these methods. We focus our analysis on the DLCP and MST approaches which yield more realistic road networks (Figure \@ref(fig:fig2)). We compared these methods to an alternative iterative least cost path method developed by Hardy [ref].

@@

## Performance Metrics

In order to assess the performance of the projections, five characteristics of the projected and observed road networks were calculated and compared: 1) Road presence, measured as whether or not roads are present in each pixel; 2) Road density, measured as density of roads in meters per hectare (m/ha); 3) Distance to nearest road, measured as distance from each raster cell to the nearest road; 4) Road disturbance footprint, measured with a 1 km buffer around roads [@ibisch2016a]; and 5) Forestry disturbance footprint, measured with a 500 m buffer around roads and cutblocks. The forestry disturbance footprint is a variant of the buffered anthropogenic disturbance metric that is used as an indicator of boreal caribou critical habitat [@johnson2020], and road density is is often used in studies of anthropogenic impacts on wildlife [@beazley2004; more]. Forestry disturbance footprint and road density were calculated using the XX and rasterizeLineDensity functions in the `caribouMetrics` R package [@hughes]. Distance to nearest road was calculated using a fast parallel moving window method implemented in the getDistFromSource function of the `roads` package. Only the road and forestry disturbance footprints were calculated for the cutblocks only method since the other characteristics are not applicable without roads. All road characteristics were stored in raster format.

@@

The aspatial performance of the projections was assessed by calculating the mean values of each road network characteristic and comparing the projected and observed values. This was done across the TSA and within cutblocks by masking the road network characteristic rasters with cutblock polygons and calculating means within cutblocks. Spatial performance was assessed in two different ways. First, difference maps were created by subtracting the projected distance to the nearest road from the observed in order to highlight discrepancies between the road networks. The observed and projected road networks were then overlain on the difference maps and regions of major discrepancies were investigated using Google satellite imagery in QGIS. Second, performance according to the three binary metrics (road presence, road disturbance footprint, and forestry disturbance footprint) was assessed by categorizing pixels of a projected raster into 5 categories compared with the observed raster: 1) pre-existing footprint; 2) agree roadless, meaning that both projected and observed show no footprint; 3) agree roaded, meaning that both projected and observed show new footprint; 4) false positive, meaning the projection shows a footprint but no footprint was observed; and 5) false negative, meaning the projection shows no footprint but a footpint was observed. These categories were summarized for each projection and characteristic and used to calculate sensitivity, precision and F-measure of each metric [@lee2021]. Sensitivity, also known as recall, gives the proportion of the true positives that were projected, while precision gives the proportion of positives projected that are true. Sensitivity will be lower when observed roads are missed and precision will be lower when roads are projected that were not observed. F-measure attempts to balance sensitivity and precision to give an overall score.

@@

<!-- Discussion pts
- disturbance footprints and distance to road are easier to project than road density and road presence.
- density of landing pts has a larger effect than choice of road projection algorithm. So maybe don't get too worked up about the differences between one algorithm and another. 
-->

@@

<!-- NOTE David's manuscript rough draft document includes other comments and bullets that might be relevant to rewriting the Intro and Discussion -->

## Sarah's thoughts

- The projections work reasonably well for strategic planning purposes, especially if we are not concerned with roads inside cutblocks.

- The differences between algorithms were fairly small and it would be more valuable to focus on the accuracy of input data such as costs and cutblock locations rather than fine tuning the algorithms.

- This example projected new roads over fairly short distances, new landings were always fairly close to existing roads or other landings so there was not a lot of space for the projection to go wrong. The results would likely be less accurate if the landings were spread further apart.

>>>>

---
title: "Assessment and implementation of resource road simulation methods for projection of cumulative effects on wildlife"
author:
  - name: Josie Hughes
    email: josie.hughes@ec.gc.ca
    institute: [ECCC]
    correspondence: true
  - name: Sarah Endicott
    email: sarah.endicott@ec.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: David Lapins
    email: david.lapins@pc.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Kyle Lochhead
    email: Kyle.Lochhead@gov.bc.ca
    institute: [FLNR]
    correspondence: false
  - name: Gregory Paradis
    email: gregory.paradis@ubc.ca
    institute: [UBC]
    correspondence: false

@@

The impacts of roads on wildlife are complex and varied. Mortality risk and stress on or near roads can be elevated due to vehicle collisions, hunting pressure, predation, noise and other stressors, but roads can also provide food, facilitate movement, and provide refuge [@hill_review_2021;@trombulak_review_2000;@fahrig_effects_2009].    
In Canadian forests, there is ample evidence of linear features attracting predators such as black bears and wolves seeking food or easy movement, while repelling prey such as caribou and moose [@dickie_corridors_2020;@polfus_identifying_2011;@mumma_predation_2018;@beauchesne_disentangling_2013;@leblond_avoidance_2013;@laurian_interactions_2012]; there is also evidence of predators avoiding features heavily used by humans [@muhly_human_2011;@whittington_towns_2022]. Changes in mortality risks, habitat quality, and connectivity associated with anthropogenic disturbance can in turn alter movement patterns [@tucker_moving_2018], demography and population viability [@johnson_science_2020; @decesare_linking_2014; @barrientos_lost_2021;@apps_spatial_2013;@pinard_calving_2012;@fahrig_effects_2009]. Roads can also facilitate spread of invasive species [@riitters_landscape_2017], and enable other development activities [@johnson_growth-inducing_2020].

Increasingly rapid environmental change, improvements in available cyberinfrastructure, and development of open source data, tools and methods have contributed to growing interest in ecological forecasting models [@bodner_bridging_2021; @dietze_ecological_2017; @mcintire_perfict_2022]. In particular, forest landscape models that can account for harvest, fire, and other disturbances are widely and increasingly used to project the cumulative effects of climate change and forest management [@petter_how_2020; @albrich_simulating_2020; @hof_editorial_2021; @tremblay_harvesting_2018;@snell_importance_2018;@daniel_incorporating_2017;@stewart_climate-informed_2023;@micheletti_assessing_2021;@leblond_there_2022; @bouderbala_long-term_2022; @barros_empowering_2023], building on a rich history of spatial-stochastic models of forest growth and succession following disturbance [@sturtevant_understanding_2021]. Despite the importance of roads for wildlife, these models generally do not include road network projections. For example, in a recent collection of articles focussed on “Using Landscape Simulation Models to Help Balance Conflicting Goals in Changing Forests” [@hof_editorial_2021], only 4 of 14 mentioned roads [@haga_scenario_2020; @leston_quantifying_2020; @mikusinski_strengthening_2021; @norris_forecasting_2021], and only one includes a road projection [@leston_quantifying_2020].

Leston et al. (2020) project forest road network development using Patchworks [@spatial_planning_systems_patchworks_nodate], then add fire. This example highlights that roads are also important for harvest planning, and that road network projection methods are integrated into a widely used optimization-based spatial harvest planning tool. However, Patchworks is proprietary closed-source commercial software, and the road network projection algorithm contained therein is difficult to assess or fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017]. Published methods intended to support strategic planning [@anderson_projecting_2004;@clark_three-stage_2000;@dean_finding_1997] require inputs that are not publicly available, and employ algorithms that are too complex and detailed for application at regional, provincial or national scales. More promising options for projecting forest road network development at large scales include an iterative least cost path method developed by Hardy [-@hardy_forest_2019], and least cost path and minimum spanning tree algorithms used to project roads and reconstruct development history in British Columbia [@lochhead_demo_2018; @lochhead_linking_2022]. The former method is available as QGIS and Landis plugins, and the latter is embedded in a larger open source modelling project that is specific to British Columbia [@muhly_castor_nodate]. None of these methods have been formally assessed, compared, published, or implemented in a way that allows for easy integration into the full variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g. @barros_empowering_2023; @daniel_state-and-transition_2016; @scheller_forest_2004].

In this work, we address the need for adequately accurate forest road network projection tools that are also widely and easily applicable by implementing simple iterative least cost path and minimum spanning tree methods in an open source R package. We compare the ability of these methods to reconstruct observed forest road network development since 1990 in a mountainous region of British Columbia using performance metrics relevant to wildlife. The adequacy of a road network projection method depends on the purpose of the projection, and it is important to select performance metrics that reflect the purpose. Common predictors in models of the impacts of roads on wildlife include distance to roads [], buffered disturbance footprint [@johnson_science_2020;@ibisch_global_2016], and density []. We expect projecting these measures to be less challenging than projecting the spatial location of each road segment, and we do not expect good projections of road locations from the algorithms considered here. Our goal is to assess the ability of simple iterative least cost path and minimum spanning tree methods to project footprint, distance and density metrics that are commonly used to predict impacts of roads on wildlife.

<!--
See notes here:
https://docs.google.com/document/d/1UQ_iC69QX0qX86NM6zSSQ036CgslN-uCXCdBY-4nU5M/edit
Effects of roads on wildlife include increased mortality from various types of encounters with humans, alterations in behaviour and interactions among predators and prey, and spread of invasive species (Trombulak and Frissel 2001). In boreal forests and coniferous forests of the Columbia and Rocky mountains, notable impacts of logging roads include facilitation of movement by wolves (Mumma et al. 2018; Dickie et al. 2020), attraction of bears to roadside food resources (Roever et al. 2010; Bastille-Rousseau et al. 2011; Dickie et al. 2020), increased mortality caused by humans near roads and trails (Roever et al. 2010), and avoidance by ungulates (Laurian et al. 2015; Dickie et al. 2020). Forest dwelling boreal woodland and mountain caribou avoid roads leading to a loss of available habitat (Polfus et al. 2011; Dickie et al. 2020), and suffer increased mortality near roads (Bastille-Rousseau et al. 2011; Mumma et al. 2018). At a range scale, survival and recruitment of boreal caribou declines with anthropogenic disturbance (Johnson et al. 2020). Eight herds of Southern Mountain Caribou have been extirpated or likely extirpated, and remaining herds are considered Endangered by COSEWIC  and listed as threatened under the Canadian Species at Risk Act (SARA) (Palm et al. 2019). Most boreal caribou populations with ranges that include intensive forest management activity are not self-sustaining (ECCC 2011; Johnson et al. 2020), and boreal caribou are also listed as threatened under SARA.

Nonlinear responses of elk to road density (Frair et al. 2008). Black bear responses to roads vary with road type (Zeller et al. 2020), and selection for roadside vegetation can increase neonatal boreal woodland caribou mortality in areas near roads . 
For more careful reading: Flyal et al. 2019; Jaeger et al. 2005; Laurance et al. 2014; Psaralexi et al. 2017; Apps et al 2014; Beyer et al. 2014; Blagdon and Johnson 2021; Suzuki and Parker 2016; Fahrig and Rytwinski 2009; Coffin et al. 2021; Crist et al. 2005; Grantham et al. 2020;Johnson et al. 2019; Meijer et al. 2018; Langeveld et al. 2009

Continue reviewing pdfs in downloads -

-->

@@

<!-- TO DO: dynamic links to figs and tables. -->
We chose the Revelstoke timber supply area (TSA 27), in British Columbia, Canada as an assessment area because forestry is the dominant resource extraction activity in this region and much of the road network was built to access to forest harvest cutblocks (Figure 1). This allows projection of the forest road network from historic forest harvest data. Approximately 55% of the area is non-forested land where road building is difficult or impossible (e.g. alpine, lakes, swamp, brush, rock) [@brown_revelstoke_2010], so the area provides opportunities to observed the responses of road projection algorithms to barriers. Revelstoke is in the interior-wet belt of the province where forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%) [@ministry_of_forests_revelstoke_2021]. The area contains wildlife species that are sensitive to roads (refs) including endangered southern mountain woodland caribou (refs).

@@

Our road network data was obtained by combining the digital road atlas (DRA) [@data_bc_digital_2021] and forest tenure road [@british_columbia_ministry_of_forests_lands_natural_resource_operations_and_rural_development__flnrord_forest_2021] data sets for BC. To avoid situations where the same road was included in both data sets with slightly different coordinates, we removed all roads from the DRA data set that were within 50 m of a road in the forest tenure road data set. The existing road network, used as the starting point for projections, was created by filtering out forest tenure roads with an award date greater than 1990. All roads from the DRA data set were assumed to be built before 1990. Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset [@government_of_british_columbia_harvested_2021]. Cutblock information is compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes. The dataset does not exclude non-harvested 'reserved' areas, and satellite image change detection may not always represent the entire cutblock area [@government_of_british_columbia_harvested_2021]. Forest harvest cutblocks with harvest years greater than 1990 were filtered and used as targets for the projections.

To spatially define the cost of building and traveling on a forestry road we developed a simple model informed by slope and major barriers (Table 1). Following the Interior Appraisal Manual of BC, we assume the base cost per kilometer of road is a linear function of slope [@government_of_british_columbia_interior_2021], and that road construction across major water bodies is prohibitively expensive (Table 1). Existing roads are assigned a cost of 0. The cost raster consists of 1703936 pixels with a resolution of 100 meters by 100 meters.

@@

All data sets were accessed using the `bcdata` R package and filtered to the extent of the Revelstoke TSA [@teucher_bcdata_2021].

@@

Simple methods for projecting low-cost routes from an existing road network to target locations (i.e. harvest cutblocks) include simple least cost paths (LCP), dynamic least cost paths (DLCP) and least cost path minimum spanning trees (MST) \@ref(fig:fig2)). We use Dijkstra's shortest path algorithm [-@dijkstra_note_1959] to determine the least cost path between a target location (e.g. centroid of a cutblock) and the existing road network. Least cost paths to each cutblock is determined independently in simple method (LCP). The dynamic least cost paths (DLCP) method evaluates the path to each cutblock in sequence so that earlier paths are considered in the cost for later paths. Sequence is determined by a heuristic such as the distance to the nearest existing road in these examples. The minimum spanning tree (MST) method calculates the minimum cost network to connecting all cutblocks to the road network using Kruskal's algorithm [-@kruskal_shortest_1956]. See Appendix A for a more detailed explanation of these methods. We focus our analysis on the DLCP and MST approaches which yield more realistic road networks (Figure \@ref(fig:fig2)). We compared these methods to an alternative iterative least cost path method developed by Hardy [ref].

@@

## Performance Metrics
In order to assess the performance of the projections, five characteristics of the projected and observed road networks were calculated and compared: 1) Road presence, measured as whether or not roads are present in each pixel; 2) Road density, measured as density of roads in meters per hectare (m/ha); 3) Distance to nearest road, measured as distance from each raster cell to the nearest road; 4) Road disturbance footprint, measured with a 1 km buffer around roads [@ibisch_global_2016]; and 5) Forestry disturbance footprint, measured with a 500 m buffer around roads and cutblocks. The forestry disturbance footprint is a variant of the buffered anthropogenic disturbance metric that is used as an indicator of boreal caribou critical habitat [@johnson_science_2020], and road density is is often used in studies of anthropogenic impacts on wildlife [@beazley_road_2004; more]. Forestry disturbance footprint and road density were calculated using the XX and rasterizeLineDensity functions in the `caribouMetrics` R package [@hughes]. Distance to nearest road was calculated using a fast parallel moving window method implemented in the getDistFromSource function of the `roads` package. Only the road and forestry disturbance footprints were calculated for the cutblocks only method since the other characteristics are not applicable without roads. All road characteristics were stored in raster format.

@@

The aspatial performance of the projections was assessed by calculating the mean values of each road network characteristic and comparing the projected and observed values. This was done across the TSA and within cutblocks by masking the road network characteristic rasters with cutblock polygons and calculating means within cutblocks. Spatial performance was assessed in two different ways. First, difference maps were created by subtracting the projected distance to the nearest road from the observed in order to highlight discrepancies between the road networks. The observed and projected road networks were then overlain on the difference maps and regions of major discrepancies were investigated using Google satellite imagery in QGIS. Second, performance according to the three binary metrics (road presence, road disturbance footprint, and forestry disturbance footprint) was assessed by categorizing pixels of a projected raster into 5 categories compared with the observed raster: 1) pre-existing footprint; 2) agree roadless, meaning that both projected and observed show no footprint; 3) agree roaded, meaning that both projected and observed show new footprint; 4) false positive, meaning the projection shows a footprint but no footprint was observed; and 5) false negative, meaning the projection shows no footprint but a footpint was observed. These categories were summarized for each projection and characteristic and used to calculate sensitivity, precision and F-measure of each metric [@lee_prediction_2021]. Sensitivity, also known as recall, gives the proportion of the true positives that were projected, while precision gives the proportion of positives projected that are true. Sensitivity will be lower when observed roads are missed and precision will be lower when roads are projected that were not observed. F-measure attempts to balance sensitivity and precision to give an overall score.

@@

@@

<!-- Discussion pts
- simple algorithms, practical to apply over large areas with minimal available information (unlike operational models)
- as expected, not great at projecting exact road locations. sources of discrepancy include switchback, stream crossings, inaccurate harvest locations. All of these could be improved with more detailed high resolution cost & harvest data.
- density of landing pts has a larger effect than choice of road projection algorithm. So maybe don't get too worked up about the differences between one algorithm and another. if more accurate projections are required, focus first on refining cost & harvest landing locations.
- Selected revelstoke because it is topographically complex, so a good challenge. However, This example projected new roads over fairly short distances, new landings were always fairly close to existing roads or other landings so there was not a lot of space for the projection to go wrong. The results would likely be less accurate if the landings were spread further apart. But also less topographically complex landscapes may be simpler to project.
- For some purposes, projections from relatively crude cost models may be adequate. For this example landscape projections of disturbance footprints and distance to road are pretty good (and better than nothing). Density is harder to project partly because none of these algorithms can project variation in density within 100 m pixels. If projection of impacts on wildlife is a goal, it may be helpful to consider including more easily projectable road predictors in wildlife response models (Bodner,). If interested in predictors that we did not consider, see example script. Can evaluate alternative metrics without rerunning the projections.
-->

<!-- TO CONSIDER: provide example script that would make it relatively easy for someone to assess the performance of a new road metric of their choice. They don't need to run the projections to do this - just summarize the outputs using a method of their choice. -->
